---
title: "Enrichment Analysis"
author: 
  - name: Juan R Gonzalez
    affiliation: 
    - Bioinformatics Research Group in Epidemiology, Barcelona Institute for Global Health, Spain
    email: juanr.gonzalez@isglobal.org
date: "July, 2019"
output:
  BiocStyle::html_document:
    toc: yes
    toc_depth: 3
  BiocStyle::pdf_document:
    fig_caption: yes
    toc_depth: 3
abstract: Overview of existing methods for enrichment analysis of gene expression data with regard to functional gene sets and pathways. Functionality for differential expression analysis, set- and network-based enrichment analysis, along with visualization and exploration of results. Introduction to concepts for enrichment analysis of genomic regions and regulatory elements. 
---

```{r style, echo = FALSE, results = 'asis'}
options(width=120)
knitr::opts_chunk$set(cache=TRUE, fig.align = TRUE, 
                      warning = FALSE,
                      message = FALSE, comment = "")
```


# Introduction 

Test whether known biological functions or processes are over-represented
(= enriched) in an experimentally-derived gene list, e.g. a list of
differentially expressed (DE) genes. See
[Goeman and Buehlmann, 2007](https://doi.org/10.1093/bioinformatics/btm051) for a critical review.


A convenient way to find DE gene sets is by detecting the so-called *functional enrichment* in two steps. First, we search for DE genes. Second, we verify whether these DE genes belong to the gene set in a proportion that exceeds any expectation of finding that number of genes in that gene set by chance alone. A straightforward way to assess this hypothesis consists of applying a hypergeometric test, which in fact corresponds to the one-tailed Fisher's exact test, following the so-called **urn model**.

A *hypergeometric test* assesses whether a number of successes in a          sequence of draws follows a hypergeometric distribution. The *hypergeometric distribution*  is a discrete probability distribution that describes the number of successes in a sequence of draws from a finite population without replacement, just as the binomialdistribution describes the number of successes for draws with replacement. In our context this involves the following quantities:

|                   |  DE |  non-DE | total |
|-------------------|:---:|:-------:|:-----:|
| Insiside Gene Set |  k  |   m-k   |   m   |
| Outside Gene Set  | n-k | N+k-n-m |  N-m  |
| total             |  n  |   N-m   |   N   |


where:

- $N$ is the total number of genes considered (gene universe).
- $n$ is the number of DE genes.
- $m$ is the number of genes in the gene set.
- $k$ is the number of DE genes in the gene set.


Given these quantities and a random variable $X$ representing the
possible outcomes of a hypergeometric process, the probability of
getting exactly $k$ genes inside a particular gene set is given by

$$\textrm{Pr}(X=k)=\frac{{m\choose k} {N-m\choose n-k}}{{N\choose n}}$$
Therefore, the probability of getting $k$ or more genes inside a
particular gene set, as a hypergeometric random process, is

$$\textrm{Pr}(X >= k)=\sum_{x=k}^n \frac{{m\choose x} {N-m\choose n-x}}{{N\choose n}}$$
This probability is the quantity being used as surrogate for the
enrichment of the functional gene set with our list of DE genes,
where the lower the value the less likely that we observe those
$k$ genes inside the given gene set by pure chance and thus the
more enriched that this gene set is enriched with our DE genes.


**Example**: Transcriptomic study, in which 12,671 genes have been tested for
differential expression between two sample conditions and 529 genes were found
DE. Among the DE genes, 28 are annotated to a specific functional gene set (or pathway), which contains in total 170 genes. This setup corresponds to a 2x2 contingency table,


```{r deTbl}
deTable <-  matrix(c(28, 142, 501, 12000),
            nrow = 2,
            dimnames = list(DE=c("yes","no"),
                            GeneSet=c("in","out")))
              
deTable
```

where the overlap of 28 genes can be assessed based on the hypergeometric distribution.
This corresponds to a one-sided version of Fisher's exact test, yielding here a
significant enrichment.

```{r fisher}
fisher.test(deTable, alternative = "greater")
```

This basic principle is at the foundation of major public and commercial enrichment
tools such as [DAVID](https://david.ncifcrf.gov/) and 
[Pathway Studio](https://www.pathwaystudio.com).

Although gene set enrichment methods have been primarily developed and applied
on transcriptomic data, they have recently been modified, extended and applied
also in other fields of genomic and biomedical research. This includes novel
approaches for functional enrichment analysis of proteomic and metabolomic data
as well as genomic regions and disease phenotypes,
[Lavallee and Yates, 2016](https://doi.org/10.1002/0471250953.bi1328s53),
[Chagoyen et al., 2016](https://doi.org/10.1007/978-1-4939-3572-7_20),
[McLean et al., 2010](https://doi.org/10.1038/nbt.1630),
[Ried et al., 2012](https://doi.org/10.1002/gepi.21617).


# Gene expression-based enrichment analysis

The first part of the workshop is largely based on the
`r Biocpkg("EnrichmentBrowser")` package, which implements an analysis pipeline
for high-throughput gene expression data as measured with microarrays and
RNA-seq. In a workflow-like manner, the package brings together a selection of
established Bioconductor packages for gene expression data analysis. It
integrates a wide range of gene set enrichment analysis methods and facilitates
combination and exploration of results across methods.

```{r loadEBrowser}
library(EnrichmentBrowser)
```

Further information can be found in the 
[vignette](http://www.bioconductor.org/packages/release/bioc/vignettes/EnrichmentBrowser/inst/doc/EnrichmentBrowser.pdf)
and [publication](https://doi.org/10.1186/s12859-016-0884-1).


## A primer on terminology, existing methods & statistical theory

**Gene sets, pathways & regulatory networks**

Gene sets are simple lists of usually functionally related genes without further
specification of relationships between genes.

Pathways can be interpreted as specific gene sets, typically representing a
group of genes that work together in a biological process. Pathways are commonly divided in metabolic and signaling pathways.
Metabolic pathways such as glycolysis represent biochemical substrate conversions
by specific enzymes. Signaling pathways such as the MAPK signaling pathway describe 
signal transduction cascades from receptor proteins to transcription factors, 
resulting in activation or inhibition of specific target genes.

Gene regulatory networks describe the interplay and effects of regulatory
factors (such as transcription factors and microRNAs) on the expression of their
target genes.

**Resources**

[GO](http://www.geneontology.org), [KEGG](http://www.genome.jp/kegg) and [MSigDB](http://software.broadinstitute.org/gsea/msigdb/collections.jsp)
annotations are most frequently used for the enrichment analysis of 
functional gene sets. Despite an increasing number of gene set and pathway 
databases, they are typically the first choice due to their long-standing 
curation and availability for a wide range of species.

*GO*: The Gene Ontology (GO) consists of three major sub-ontologies that classify
gene products according to molecular function (MF), biological process (BP) and 
cellular component (CC). Each ontology consists of GO terms that define MFs, BPs
or CCs to which specific genes are annotated. The terms are organized in a
directed acyclic graph, where edges between the terms represent
relationships of different types. They relate the terms according to a parent-child
scheme, i.e. parent terms denote more general entities, whereas child terms represent
more specific entities.

*KEGG*: The Kyoto Encyclopedia of Genes and Genomes (KEGG) is a collection of 
manually drawn pathway maps representing molecular interaction and reaction networks.
These pathways cover a wide range of biochemical processes that can be divided in
7 broad categories: metabolism, genetic and environmental information processing,
cellular processes, organismal systems, human diseases, and drug development. 
Metabolism and drug development pathways differ from pathways of the other 5
categories by illustrating reactions between chemical compounds.
Pathways of the other 5 categories illustrate molecular interactions between
genes and gene products. 

*MSigDB*: Molecular Signatures Database (MSigDB) are divided into 8 major collections, and several sub-collections. These include: 

  - *H: hallmark gene sets*: Hallmark gene sets summarize and represent specific well-defined biological states or processes and display coherent expression.

  - *C1: positional gene sets*: Gene sets corresponding to each human chromosome and each cytogenetic band that has at least one gene.
  
  - *C2: curated gene sets*: Gene sets curated from various sources such as online pathway databases, the biomedical literature, and knowledge of domain experts. It contains 5 sub-collections.

  - *C3 motif gene sets*: Gene sets representing potential targets of regulation by transcription factors or microRNAs. It has 2 sub-collections.
  
  - *C4 computational gene sets*: Computational gene sets defined by mining large collections of cancer-oriented microarray data. It has 4 sub-collections. 
  
  - *C5 GO gene sets*: Gene sets that contain genes annotated by the same GO term. The C5 collection is divided into three sub-collections based on GO ontologies: BP, CC, and MF.
  
  - *C6 oncogenic signatures*: Gene sets that represent signatures of cellular pathways which are often dis-regulated in cancer.
  
  - *C7 inmunologic signatures*: Gene sets that represent cell states and perturbations within the immune system.
  
  
**Statistical approaches**

The two predominantly used enrichment methods are:

- Overrepresentation analysis (ORA), testing whether a gene set contains 
disproportional many genes of significant expression change, based on the 
procedure outlined in the first section
- Gene set enrichment analysis (GSEA), testing whether genes of a gene set 
accumulate at the top or bottom of the full gene vector ordered by direction
and magnitude of expression change
[Subramanian et al., 2005](https://doi.org/10.1073/pnas.0506580102). However, the term *gene set enrichment analysis* nowadays subsumes a general 
strategy implemented by a wide range of methods 
[Huang et al., 2009](https://doi.org/10.1093/nar/gkn923). 
Those methods have in common the same goal, although approach and statistical
model can vary substantially
[Goeman and Buehlmann, 2007](https://doi.org/10.1093/bioinformatics/btm051),
[Khatri et al., 2012](https://doi.org/10.1371/journal.pcbi.1002375).

To better distinguish from the specific method, some authors use the term
*gene set analysis* to denote the general strategy.
However, there is also a specific method from
[Efron and Tibshirani, 2007](https://doi.org/10.1214/07-AOAS101) of this name.

Goeman and Buehlmann further raise several critical issues concerning the 2x2 ORA:

- rather arbitrary classification of genes in DE / not DE
- based on gene sampling, although sampling of subjects is appropriate 
- unrealistic independence assumption between genes, resulting in highly
anti-conservative *p*-values

With regard to these statistical concerns, GSEA is considered superior:

- takes all measured genes into account 
- subject sampling via permutation of class labels  
- the incorporated permutation procedure implicitly accounts for correlations
between genes

However, the simplicity and general applicability of ORA is unmet by subsequent
methods improving on these issues. For instance, GSEA requires the expression data
as input, which is not available for gene lists derived from other experiment types.
On the other hand, the involved sample permutation procedure has been proven 
inaccurate and time-consuming
[Efron and Tibshirani, 2007](https://doi.org/10.1214/07-AOAS101),
[Phipson and Smyth, 2010](https://doi.org/10.2202/1544-6115.1585),
[Larson and Owen, 2015](https://doi.org/10.1186/s12859-015-0571-7).


[Khatri et al., 2012](https://doi.org/10.1371/journal.pcbi.1002375) have taken a
slightly different approach by classifying methods along the timeline of 
development into three generations:

1. Generation: ORA methods based on the 2x2 contingency table test,
2. Generation: functional class scoring (FCS) methods such as GSEA, which compute
gene set (= functional class) scores by summarizing per-gene DE statistics,
3. Generation: topology-based methods, explicitly taking into account interactions
between genes as defined in signaling pathways and gene regulatory networks
([Geistlinger et al., 2011](https://doi.org/10.1093/bioinformatics/btr228) for an example).

Although topology-based (also: network-based) methods appear to be most realistic,
their straightforward application can be impaired by features that are not-detectable
on the transcriptional level (such as protein-protein interactions) and insufficient network knowledge
[Geistlinger et al., 2013](https://doi.org/10.1093/nar/gkt631),
[Bayerlova et al., 2015](https://doi.org/10.1186/s12859-015-0751-5).

Given the individual benefits and limitations of existing methods,
cautious interpretation of results is required to derive valid conclusions. 
Whereas no single method is best suited for all application scenarios, applying multiple methods can be beneficial.
This has been shown to filter out spurious hits of individual methods, thereby reducing the outcome to gene sets accumulating evidence from different methods
[Geistlinger et al., 2016](https://doi.org/10.1186/s12859-016-0884-1),
[Alhamdoosh et al., 2017](https://doi.org/10.1093/bioinformatics/btw623).


## Basic analysis using `GOstats`

Let us illustrate how to perform ORA (enrichment analysis) using microarray data. First, we need to perform DE analysis. Our interest is to compare gene expression between cell lines that have been treated with dexamethasone or not.

```{r pdataAirway}
library(airway)
data(airway)
table(airway$dex)
``` 

Then, let us analyzed the RNA-seq count data using `edgeR` that uses generalized linear models (negative binomial) (`DESeq2`, `voom` or any other method could be used instead). NOTE that we avoid normalization step which is normally required (`DESeq2` incorporates this step in its pipeline).

```{r DE_airway}
library(edgeR)
airCounts <- assay(airway)
group <- airway$dex
d <- DGEList(counts=airCounts, group=group)
design <- model.matrix( ~ group)
d <- estimateDisp(d, design)
d$common.dispersion

# LRT test
fit <- glmFit(d, design)
lrt <- glmLRT(fit,coef=2)
topTags(lrt)
```

The latter value is the coefficient of variation that tells us the
proportion by which true expression levels may vary between replicate
samples. There is very little extra variability between replicates in
these data. The common dispersion NB model is going to be effectively
Poisson.

We call DE genes with a minimum 2-fold change of expression at
a maximum FDR of 1%:

```{r de_search}
tt <- topTags(lrt, n=Inf)
mask <- tt$table$FDR < 0.01 &
        abs(tt$table$logFC) > log2(2)
deGenes <- rownames(tt$table[mask, ])
head(deGenes)
length(deGenes)
```

The gene univese is obtained by:

```{r gene_universe}
geneUniverse <- rownames(tt$table)
length(geneUniverse)
```

In order to asses  functional enrichment, both DE gene list and gene universe must be annotated in Entrez IDs:

```{r genes_to_entrez}
library(org.Hs.eg.db)
deGenes <- unlist(mget(deGenes, envir=org.Hs.egENSEMBL2EG,
                       ifnotfound = NA))

geneUniverse <- unlist(mget(geneUniverse, envir=org.Hs.egENSEMBL2EG,
                       ifnotfound = NA))
``` 

Now, the enrichment analysis for GO is performed by

```{r functional_air}
library(GOstats)
params <- new("GOHyperGParams", geneIds=deGenes,
              universeGeneIds=geneUniverse,
              annotation="org.Hs.eg.db", ontology="BP",
              pvalueCutoff=0.05, conditional=FALSE,
              testDirection="over")

hgOver <- hyperGTest(params)
hgOver
```

The table can be generated by

```{r summary_hyperG_test_over}
head(summary(hgOver))
```


We can alos explore the results by generating an html report:

```{r report_hyperG_test_over, eval=FALSE}
htmlReport(hgOver, file="goBPuncond.html")
```

The KEGG enrichment is performed by:

```{r kegg_enrich}
library(KEGG.db)
params.kegg <- new("KEGGHyperGParams", geneIds=deGenes,
                   universeGeneIds=geneUniverse,
                   annotation="org.Hs.eg.db", 
                   pvalueCutoff=0.05, 
                   testDirection="over")

hgOver.kegg <- hyperGTest(params.kegg)
head(summary(hgOver.kegg))
```

## Analysis using `clusterProfiler` 

[clusterProfiler]() performs statistical analysis and visualization of functional profiles for genes and gene clusters. It includes GO, KEEG and a general function to perform ORA for any database such as MSigDB or [DisGeNET](http://www.disgenet.org/) among others.

```{r GO_clusterProfiler}
library(clusterProfiler)
ans.go <- enrichGO(gene = deGenes, ont = "BP",
                   OrgDb ="org.Hs.eg.db",
                   readable=TRUE,
                   pvalueCutoff = 0.05)
tab.go <- as.data.frame(ans.go)
tab.go<- subset(tab.go, Count>5)
tab.go[1:5, 1:6]
```


```{r KEGG_clusterProfiler}
ans.kegg <- enrichKEGG(gene = deGenes,
                       organism = 'hsa',
                       pvalueCutoff = 0.05)
tab.kegg <- as.data.frame(ans.kegg)
tab.kegg<- subset(tab.kegg, Count>5)
tab.kegg[1:5, 1:6]
```


For external databases such as DisGeNET or MSigDB collections, databases must be imported into R. This is how DisGeNET is imported into R

```{r disgenet_import}
gda <- read.delim("data/curated_gene_disease_associations.tsv.gz")
disease2gene <- gda[, c("diseaseId", "geneId")]
disease2name <- gda[, c("diseaseId", "diseaseName")]
```


Then, the enrichment is perform using `enricher` function:

```{r disgenet_clusterProfiler}
ans.dis <- enricher(deGenes, TERM2GENE=disease2gene,
                    TERM2NAME=disease2name)
tab.dis <- as.data.frame(ans.dis)
tab.dis<- subset(tab.dis, Count>5)
tab.dis[,1:6]
```


This is the R code for transcription factors which is one of the sub-collection of C3 available at  [MSigDb](http://software.broadinstitute.org/gsea/msigdb) for
additional gene set collections.


```{r c3_tf}
c3.tf <- read.gmt("c:/Juan/CREAL/HELIX/pathways/GSEA/c3.tft.v6.2.entrez.gmt")

ans.tf <- enricher(deGenes, TERM2GENE=c3.tf)
tab.tf <- as.data.frame(ans.tf)
tab.tf<- subset(tab.tf, Count>5)
tab.tf[1:5,1:5]
```

# Genomic region enrichment analysis

Microarrays and next-generation sequencing are also widely applied for
large-scale detection of variable and regulatory genomic regions, e.g. single
nucleotide polymorphisms, copy number variations, and transcription factor
binding sites. Such experimentally-derived genomic region sets are raising similar questions
regarding functional enrichment as in gene expression data analysis. Of particular interest is thereby whether experimentally-derived regions
overlap more (enrichment) or less (depletion) than expected by chance with
regions representing known functional features such as genes or
promoters. 

The `regioneR` package implements a general framework for testing
overlaps of genomic regions based on permutation sampling.
This allows to repeatedly sample random regions from the genome, matching size
and chromosomal distribution of the region set under study.
By recomputing the overlap with the functional features in each permutation,
statistical significance of the observed overlap can be assessed. To demonstrate the basic functionality of the package, we consider the analysis of copy number variants (CNV). 


Let us assume, we want to assess the overlap between the significant CNV regions obtained after comparing normal vs tumor samples from  [TCGA](https://www.cancer.gov/about-nci/organization/ccg/research/structural-genomics/tcga) and protein-coding regions. We now recall the results we obtained in the association analysis

```{r oad_brca_sig}
load("data/brcaSigCNV.Rdata")
brca.gr.sig
```

To obtain the location of protein-coding genes, we query available Human annotation from Ensembl

```{r load_hub, eval=FALSE}
library(AnnotationHub)
ah <- AnnotationHub()
ahDb <- query(ah, pattern = c("Homo sapiens", "EnsDb"))
ahDb
```

and retrieve gene coordinates in the UMD3.1 assembly (Ensembl 92).

```{r get_coord, eval=FALSE}
ahEdb <- ahDb[["AH60977"]]
hg.genes <- genes(ahEdb)
```


We then need to verify that the `seqlevels` annotation of both `GenomicRanges` are the same:

```{r load_hg_genes, echo=FALSE}
load("data/hgGenes.Rdata")
```

```{r show_seq_names}
seqnames(hg.genes)
seqnames(brca.gr.sig)
```

otherwise we need to re-assign the `seqlevels`

```{r get_genes, eval=FALSE}
seqlevels(hg.genes) <- paste0("chr", seqlevels(hg.genes))
```

Let us run the enrichment in chromosome 1:

```{r sel_chr1}
sel.protein <- hg.genes[seqnames(hg.genes)%in%c("chr1")]
sel.protein <- sel.protein[sel.protein$gene_biotype ==
                             "protein_coding"]
sel.cnvs <- brca.gr.sig[seqnames(brca.gr.sig)%in%c("chr1")]
length(sel.cnvs)
```

We then apply an overlap permutation test with 100 permutations (`ntimes=100`), while maintaining chromosomal distribution of the CNV region set (`per.chromosome=TRUE`). Setting the option `count.once=TRUE` counts an overlapping CNV region only once. To randomly sample  regions from the entire genome no masking is required (`mask=NA`) and to draw robust conclusions a minimum of 1000 permutations should be performed


```{r test_cnvs_enrich}
library(regioneR)
library(BSgenome.Hsapiens.UCSC.hg19.masked)
res.prot <- overlapPermTest(A=sel.cnvs, B=sel.protein, 
                            mask=NA, genome="hg19",
                            ntimes=100,
                            per.chromosome=TRUE,
                            count.once=TRUE)
res.prot
```

The permutation $P$-value indicates a statistically significant overlaps of CNVs with protein coding regions. From 234 CNVs, 220 overlap with at least one protein coding gene. We can also repeat the analysis for promoters from UCSC defined as -2000 to +200 base pairs to the TSS according to the genome assembly hg19. Analyzing chromosome 1

```{r enrich_promoters_get, eval=FALSE}
file.prom <- "http://gattaca.imppc.org/regioner/data/UCSC.promoters.hg19.bed" 
promoters <- toGRanges(file.prom)
sel.prom <- promoters[seqnames(promoters) %in% c("chr1")]
```

```{r enrich_promoters_get2, echo=FALSE}
load("data/selProm.Rdata")
```

we obtain
```{r enrich_promoters}
res.prom <- overlapPermTest(sel.cnvs, sel.prom, 
                            ntimes=100, genome="hg19", 
                            count.once=TRUE)
res.prom
```

We also observe significant overlap of CNVs in regions with promoters. Notice that the function  `permTest` allows  user-defined functions for randomizing control and for evaluating additional genomic measures.
 

Note #1: The function `regioneR::permTest` allows to incorporate user-defined
functions for randomizing regions and evaluating additional measures of overlap
such as total genomic size in bp.

Note #2: The `r Biocpkg("LOLA")` package implements a genomic region ORA, which
assesses genomic region overlap based on the hypergeometric distribution using
a library of pre-defined functional region sets.

--------------------------------------------------

**EXERCISE**: To be supplied


--------------------------------------------------
--------------------------------------------------

**EXERCISE**: The file `data_exercises/GSE18123.Rdata` contains data corresponing to the GEO accesion number GSE18123. This dataset was used to test whether peripheral blood gene expression profiles could be used as a molecular diagnostic tool for distinguishing children with Autistic Spectrum Disorder (ASD) from controls by performing peripheral blood gene expression profiling on 170 patients with ASD and 115 controls collected from Boston area hospitals.

1. Determine those genes that are differentially expressed between CONTROLs and AUTISM (variable `group`).

2. Perform an enrichment analysis of GO, KEGG and DisGeNET databases using `clusterProfiler`.

3. Investigate whether the DE genes are entiched in protein-coding genes.



```{r sessionInfo}
sessionInfo()
```

[EnrichmentBrowser]: https://bioconductor.org/packages/EnrichmentBrowser
[ALL]: https://bioconductor.org/packages/ALL
[hgu95av2.db]: https://bioconductor.org/packages/hgu95av2.db
[airway]: https://bioconductor.org/packages/airway
[regioneR]: https://bioconductor.org/packages/regioneR
[BSgenome.Hsapiens.UCSC.hg19.masked]: https://bioconductor.org/packages/BSgenome.Hsapiens.UCSC.hg19.masked
[MultiAssayExperiment]: https://bioconductor.org/packages/MultiAssayExperiment
[readr]: https://cran.r-project.org/package=readr
[mogsa]: https://bioconductor.org/packages/mogsa
[limma]: https://bioconductor.org/packages/limma
[BiocStyle]: https://bioconductor.org/packages/BiocStyle
[knitr]: https://cran.r-project.org/package=knitr
[rmarkdown]: https://cran.r-project.org/package=rmarkdown
